---
title: "calculate_wGRR"
output: html_document
date: "2026-03-03"
---

```{r}
library(dplyr)
library(readr)
library(stringr)
library(ggplot2)
```

Load the raw Diamond output:

```{r setup, include=FALSE}
hits <- read_tsv('output/diamond_out.tsv',
  col_names=c("qseqid","sseqid","pident","length","qlen",
              "slen","qcovhsp","evalue","bitscore"),
  col_types="ccdddddddd", comment="#")

```

Parse the plasmid names for both query and subject:

```{r}
hits <- hits %>% mutate(
  q_plasmid = str_extract(qseqid, "[A-Z]{1,3}_?[A-Z0-9]+\\.[0-9]+"),
  s_plasmid = str_extract(sseqid, "[A-Z]{1,3}_?[A-Z0-9]+\\.[0-9]+"))

all_plasmids <- sort(unique(c(hits$q_plasmid, hits$s_plasmid)))
message("  ", length(all_plasmids), " plasmids detected")  
```

In case we want to filter low identity or coverage hits, we can adapt the following filtering script:

```{r}
message("  ", nrow(hits), " hits before filtering")

hits <- hits %>%
  filter(pident >= 30, qcovhsp >= 50,
         q_plasmid != s_plasmid)

message("  ", nrow(hits), " hits after filtering")
```

Now, nothing is filtered since we already specified these pident and qcovhsp filterings in the initial diamond call.

For each gene, keep the single best hit per subject plasmid:

```{r}
best <- hits %>% 
  group_by(qseqid, s_plasmid) %>%
  slice_max(bitscore, n=1, with_ties=FALSE) %>%
  ungroup()
```

Here we find reciprocal best hits between all proteins. In other words, if protein A hits protein B as best hit, B should also hit A as best hit to be called a RBH.

```{r}
fwd <- best %>% select(gA=qseqid, gB=sseqid, pA=q_plasmid, pB=s_plasmid, pident_fwd=pident)
rev <- best %>% select(gB=qseqid, gA=sseqid, pB=q_plasmid, pA=s_plasmid, pident_rev=pident)

rbh <- inner_join(fwd, rev, by=c("gA","gB","pA","pB")) %>%
  mutate(pident_mean = (pident_fwd + pident_rev) / 2)
```

This is Claude generated code and it seems to work pretty well, but i need to look at this. In the end, the wGRR is calculated between all plasmid pairs

```{r}
# ── 5. Gene counts per plasmid (denominator) ──────────────────────────────────
gene_counts <- hits %>%
  select(gene=qseqid, plasmid=q_plasmid) %>%
  bind_rows(hits %>% select(gene=sseqid, plasmid=s_plasmid)) %>%
  distinct() %>%
  count(plasmid, name="n_genes")

# ── 6. wGRR ───────────────────────────────────────────────────────────────────
# wGRR(A,B) = sum(pident/100 per RBH) / max(nGenes_A, nGenes_B)
message("Calculating wGRR...")
wgrr_pairs <- rbh %>%
  group_by(pA, pB) %>%
  summarise(n_rbh=n(), sum_pident=sum(pident_mean/100), .groups="drop") %>%
  left_join(gene_counts %>% rename(pA=plasmid, nA=n_genes), by="pA") %>%
  left_join(gene_counts %>% rename(pB=plasmid, nB=n_genes), by="pB") %>%
  mutate(denom=pmin(nA,nB),
         wGRR=pmin(pmax(sum_pident/denom, 0), 1))
```

Turn into symmetric wGRR matrix. First we geneate an matrix filled with 0.0 values: this is the default for pA - pB pairs that did not have a single blast hit between them. Then we fill the diagonal with 1.0: that is pA == pB. Then we add the wGRR values prepared before.

```{r}
mat <- matrix(0.0, nrow=length(all_plasmids), ncol=length(all_plasmids),
              dimnames=list(all_plasmids, all_plasmids))
diag(mat) <- 1.0
for (i in seq_len(nrow(wgrr_pairs))) {
  pA <- wgrr_pairs$pA[i]; pB <- wgrr_pairs$pB[i]; v <- wgrr_pairs$wGRR[i]
  mat[pA,pB] <- v; mat[pB,pA] <- v
}
```

Display a summary of the matrix:

```{r}
off <- mat[lower.tri(mat)]
message("\n--- wGRR summary ---")
message("  Pairs total   : ", length(off))
message("  wGRR > 0      : ", sum(off > 0))
message("  Mean wGRR     : ", round(mean(off), 4))
message("  Max  wGRR     : ", round(max(off),  4))
```
Now we turn the wGRR matrix into a distance matrix (distance = 1 - wGRR)

```{r}
dist_mat        <- 1 - mat
dist_mat[dist_mat < 0] <- 0                        # floating-point guard
diag(dist_mat)  <- 0                               # self-distance = 0

dist_obj <- as.dist(dist_mat)
```

Visualizing a distance matrix can be done in many ways. Here we explore a dimensionality reduction method (MDS) to visualize the plasmids in 2D space (somewhat similar to PCA):

```{r}
mds_result <- cmdscale(
  d   = dist_obj,
  k   = 2,
  eig = TRUE,
  add = TRUE
  )
```

Just like a PCA we can find the variance explained by the two axes.

```{r}
variance_explained <- mds_result$eig / sum(mds_result$eig[mds_result$eig > 0]) * 100
cat(round(variance_explained[1:2], 1), "%")
```
Oof that's very low! Keep that in mind while looking at the following scatter plots.

To plot a scatter plot first we get the coordinates from the mds_results

```{r}
coords <- as.data.frame(mds_result$points)
colnames(coords) <- paste0("MDS", seq_len(2))
coords$plasmid_id <- rownames(mat)
coords <- coords[, c("plasmid_id", paste0("MDS", seq_len(2)))]  
```

Plot plasmids in MDS space:

```{r}
coords %>% ggplot(aes(x = MDS1, y = MDS2)) + geom_point(alpha = 0.5) + theme_bw()
```
```{r}
metadata <- read_tsv('data/metadata.tsv', col_names = c("accession", "species", "taxid", "length"))

metadata <- metadata %>% mutate(species_clean = stringr::word(species, 1, 2))
```
Merge metadata with the plasmid coordinates, plot with color for species:

```{r}
coords <- merge(coords, metadata, by.x = "plasmid_id", by.y = "accession")
```

```{r}
coords %>% ggplot(aes(x = MDS1, y = MDS2, colour = species_clean)) + geom_point(alpha = 0.5) + theme_bw()
```
This is quite cool! It shows different clusters of plasmids, with different colors within clusters. This could indicate plasmid (parts) being shared between different species, however, we need to keep in mind the low percentage of variance explained with the MDA.

```{r}
coords %>% filter(MDS1 > 0.5) %>% filter(MDS2 > 0) %>% 
  ggplot(aes(x = MDS1, y = MDS2, colour = species_clean, size = length)) + 
  geom_point(alpha = 1) + 
  scale_colour_brewer(palette = "Paired") +
  theme_void()
 

```

```{r}
library(gghighlight)

coords %>% ggplot(aes(x = MDS1, y = MDS2, colour = species_clean)) + 
  geom_point(alpha = 0.5) + 
  gghighlight() +
  facet_wrap(~ species_clean) + 
  theme_void() 
```
I want to find plasmid pairs of plasmids from different species, and then look for high wGRR values

```{r}
metadata_minimal <- metadata %>% select(accession, species_clean)

wgrr_pairs_species <- wgrr_pairs %>% 
  merge(metadata_minimal, by.x = "pA", by.y = "accession") %>% 
  mutate(species_pA = species_clean) %>% 
  select(-species_clean) %>% 
  merge(metadata_minimal, by.x = "pB", by.y = "accession") %>% 
  mutate(species_pB = species_clean) %>% 
  select(-species_clean) %>% 
  mutate(species_pair = paste(species_pA, species_pB, sep = "-"))
```

```{r}
wgrr_pairs_species %>% group_by(species_pair) %>% 
  summarise(mean = mean(wGRR),
            median = median(wGRR), 
            min = min(wGRR),
            max = max(wGRR))
```

```{r}
wgrr_pairs_species %>% 
  filter(species_pA != species_pB) %>% 
  filter(wGRR > 0.95)
```

The final one is a really interesting result. It should cross-species plasmid pairs with very high wGRR values.


```{r}
library(ggraph)
library(tidygraph)
```
Alternative visualization in 2D space is a network (see Eduardo Rocha's papers). In these networks each plasmid becomes a node (a dot) and each wGRR connection (based on a threshold) is an edge (a line). It makes sense to only consider edges representing a certain level of similarity, e.g. wGRR > 0.5. 

```{r}
edges <- wgrr_pairs_species %>%
  filter(wGRR >= 0.3) %>%
  select(pA, pB, wGRR, species_pA, species_pB)

# Node table with species annotation
nodes <- data.frame(name = unique(c(edges$pA, edges$pB))) %>%
  left_join(
    wgrr_pairs_species %>% select(pA, species_pA) %>% distinct() %>% rename(name = pA, species = species_pA),
    by = "name"
  )

g <- tbl_graph(nodes = nodes, edges = edges, directed = FALSE, node_key = "name")
```

```{r}
ggraph(g, layout = "fr") +  # Fruchterman-Reingold, good for clusters
  geom_edge_link(aes(alpha = wGRR), colour = "grey60") +
  geom_node_point(aes(colour = species), size = 1) +
  # geom_node_text(aes(label = name), repel = TRUE, size = 2.5) +
  #scale_edge_width(range = c(0.3, 2)) +
  scale_edge_alpha(range = c(0.3, 0.9)) +
  #scale_colour_brewer(palette = "Set2") +
  theme_graph() +
  labs(title = "Plasmid relatedness network (wGRR ≥ 0.4)",
       colour = "Host species", edge_width = "wGRR") + theme_void()
```
```{r}
g <- g %>%
  activate(edges) %>%
  mutate(cross_species = species_pA != species_pB)

ggraph(g, layout = "fr") +
  geom_edge_link(aes(colour = cross_species, width = wGRR, alpha = wGRR)) +
  geom_node_point(aes(colour = species), size = 3) +
  scale_edge_colour_manual(values = c("FALSE" = "grey80", "TRUE" = "firebrick")) +
  scale_edge_width(range = c(0.3, 2)) +
  scale_edge_alpha(range = c(0.2, 0.9)) +
  theme_graph() +
  labs(title = "Cross-species plasmid transfer network",
       edge_colour = "Cross-species")
```
Repeat but only include cross species interactions in building the network in the first place

```{r}
# keep only cross-species edges
edges_cross <- wgrr_pairs_species %>%
  filter(wGRR >= 0.8, species_pA != species_pB) %>%
  select(pA, pB, wGRR, species_pA, species_pB)

# nodes automatically limited to only those in cross-species edges
nodes_cross <- data.frame(name = unique(c(edges_cross$pA, edges_cross$pB))) %>%
  left_join(
    wgrr_pairs_species %>% select(pA, species_pA) %>% distinct() %>% rename(name = pA, species = species_pA),
    by = "name"
  )

g_cross <- tbl_graph(nodes = nodes_cross, edges = edges_cross, directed = FALSE, node_key = "name")
```

```{r}
ggraph(g_cross, layout = "fr") +
  geom_edge_link(colour = "firebrick") +
  geom_node_point(aes(colour = species), size = 1) +
  # scale_colour_brewer(palette = "Set2") +
  theme_graph() +
  labs(title = "Cross-species plasmid transfer network (wGRR ≥ 0.9)",
       colour = "Host species")
```


```{r}
ggraph(g_cross, layout = 'linear', circular = TRUE) +
  geom_node_point(aes(colour = species), size = 1) +
  geom_edge_arc(linewidth = 0.2, alpha = 0.1) + 
  coord_fixed()
```
Final visualization option: UMAP

```{r}
library(uwot)
library(ggplot2)

dist_mat_object <- dist(dist_mat)

# run UMAP on the distance matrix
umap_res <- umap(dist_mat, n_neighbors = 10, min_dist = 2)

# build a plottable dataframe
umap_df <- data.frame(
  UMAP1 = umap_res[, 1],
  UMAP2 = umap_res[, 2],
  plasmid = rownames(dist_mat)
) %>%
  left_join(
    wgrr_pairs_species %>% select(pA, species_pA) %>% distinct() %>% rename(plasmid = pA, species = species_pA),
    by = "plasmid"
  )

ggplot(umap_df, aes(UMAP1, UMAP2, colour = species)) +
  geom_point(size = 1.5, alpha = 0.7) +
  theme_void() +
  labs(colour = "Host species")
```



