---
title: "calculate_wGRR"
output: html_document
date: "2026-03-03"
---

```{r}
library(dplyr)
library(readr)
library(stringr)
```

Load the raw Diamond output:

```{r setup, include=FALSE}
hits <- read_tsv('output/diamond_out.tsv',
  col_names=c("qseqid","sseqid","pident","length","qlen",
              "slen","qcovhsp","evalue","bitscore"),
  col_types="ccdddddddd", comment="#")

```

Parse the plasmid names for both query and subject:

```{r}
hits <- hits %>% mutate(
  q_plasmid = str_extract(qseqid, "[A-Z]{1,3}_?[A-Z0-9]+\\.[0-9]+"),
  s_plasmid = str_extract(sseqid, "[A-Z]{1,3}_?[A-Z0-9]+\\.[0-9]+"))
  
```

In case we want to filter low identity or coverage hits, we can adapt the following filtering script:

```{r}
message("  ", nrow(hits), " hits before filtering")

hits <- hits %>%
  filter(pident >= 30, qcovhsp >= 50,
         q_plasmid != s_plasmid)

message("  ", nrow(hits), " hits after filtering")
```

Now, nothing is filtered since we already specified these pident and qcovhsp filterings in the initial diamond call.

For each gene, keep the single best hit per subject plasmid:

```{r}
best <- hits %>% 
  group_by(qseqid, s_plasmid) %>%
  slice_max(bitscore, n=1, with_ties=FALSE) %>%
  ungroup()
```

Claude generated this part but it's not entirely clear to me what this does:

```{r}
fwd <- best %>% select(gA=qseqid, gB=sseqid, pA=q_plasmid, pB=s_plasmid, pident_fwd=pident)
rev <- best %>% select(gB=qseqid, gA=sseqid, pB=q_plasmid, pA=s_plasmid, pident_rev=pident)

rbh <- inner_join(fwd, rev, by=c("gA","gB","pA","pB")) %>%
  mutate(pident_mean = (pident_fwd + pident_rev) / 2)
```

This is also Claude and it seems to work pretty well, but i need to look at this:

```{r}
# ── 5. Gene counts per plasmid (denominator) ──────────────────────────────────
gene_counts <- hits %>%
  select(gene=qseqid, plasmid=q_plasmid) %>%
  bind_rows(hits %>% select(gene=sseqid, plasmid=s_plasmid)) %>%
  distinct() %>%
  count(plasmid, name="n_genes")

# ── 6. wGRR ───────────────────────────────────────────────────────────────────
# wGRR(A,B) = sum(pident/100 per RBH) / max(nGenes_A, nGenes_B)
message("Calculating wGRR...")
wgrr_pairs <- rbh %>%
  group_by(pA, pB) %>%
  summarise(n_rbh=n(), sum_pident=sum(pident_mean/100), .groups="drop") %>%
  left_join(gene_counts %>% rename(pA=plasmid, nA=n_genes), by="pA") %>%
  left_join(gene_counts %>% rename(pB=plasmid, nB=n_genes), by="pB") %>%
  mutate(denom=pmax(nA,nB),
         wGRR=pmin(pmax(sum_pident/denom, 0), 1))
```

