---
title: "calculate_wGRR"
output: html_document
date: "2026-03-03"
---

```{r}
library(dplyr)
library(readr)
library(stringr)
library(ggplot2)
```

Load the raw Diamond output:

```{r setup, include=FALSE}
hits <- read_tsv('output/diamond_out.tsv',
  col_names=c("qseqid","sseqid","pident","length","qlen",
              "slen","qcovhsp","evalue","bitscore"),
  col_types="ccdddddddd", comment="#")

```

Parse the plasmid names for both query and subject:

```{r}
hits <- hits %>% mutate(
  q_plasmid = str_extract(qseqid, "[A-Z]{1,3}_?[A-Z0-9]+\\.[0-9]+"),
  s_plasmid = str_extract(sseqid, "[A-Z]{1,3}_?[A-Z0-9]+\\.[0-9]+"))


all_plasmids <- sort(unique(c(hits$q_plasmid, hits$s_plasmid)))
message("  ", length(all_plasmids), " plasmids detected")  
```

In case we want to filter low identity or coverage hits, we can adapt the following filtering script:

```{r}
message("  ", nrow(hits), " hits before filtering")

hits <- hits %>%
  filter(pident >= 30, qcovhsp >= 50,
         q_plasmid != s_plasmid)

message("  ", nrow(hits), " hits after filtering")
```

Now, nothing is filtered since we already specified these pident and qcovhsp filterings in the initial diamond call.

For each gene, keep the single best hit per subject plasmid:

```{r}
best <- hits %>% 
  group_by(qseqid, s_plasmid) %>%
  slice_max(bitscore, n=1, with_ties=FALSE) %>%
  ungroup()
```

Claude generated this part but it's not entirely clear to me what this does:

```{r}
fwd <- best %>% select(gA=qseqid, gB=sseqid, pA=q_plasmid, pB=s_plasmid, pident_fwd=pident)
rev <- best %>% select(gB=qseqid, gA=sseqid, pB=q_plasmid, pA=s_plasmid, pident_rev=pident)

rbh <- inner_join(fwd, rev, by=c("gA","gB","pA","pB")) %>%
  mutate(pident_mean = (pident_fwd + pident_rev) / 2)
```

This is also Claude and it seems to work pretty well, but i need to look at this:

```{r}
# ── 5. Gene counts per plasmid (denominator) ──────────────────────────────────
gene_counts <- hits %>%
  select(gene=qseqid, plasmid=q_plasmid) %>%
  bind_rows(hits %>% select(gene=sseqid, plasmid=s_plasmid)) %>%
  distinct() %>%
  count(plasmid, name="n_genes")

# ── 6. wGRR ───────────────────────────────────────────────────────────────────
# wGRR(A,B) = sum(pident/100 per RBH) / max(nGenes_A, nGenes_B)
message("Calculating wGRR...")
wgrr_pairs <- rbh %>%
  group_by(pA, pB) %>%
  summarise(n_rbh=n(), sum_pident=sum(pident_mean/100), .groups="drop") %>%
  left_join(gene_counts %>% rename(pA=plasmid, nA=n_genes), by="pA") %>%
  left_join(gene_counts %>% rename(pB=plasmid, nB=n_genes), by="pB") %>%
  mutate(denom=pmax(nA,nB),
         wGRR=pmin(pmax(sum_pident/denom, 0), 1))
```

Turn into symmetric wGRR matrix. First we geneate an matrix filled with 0.0 values: this is the default for pA - pB pairs that did not have a single blast hit between them. Then we fill the diagonal with 1.0: that is pA == pB. Then we add the wGRR values prepared before.

```{r}
mat <- matrix(0.0, nrow=length(all_plasmids), ncol=length(all_plasmids),
              dimnames=list(all_plasmids, all_plasmids))
diag(mat) <- 1.0
for (i in seq_len(nrow(wgrr_pairs))) {
  pA <- wgrr_pairs$pA[i]; pB <- wgrr_pairs$pB[i]; v <- wgrr_pairs$wGRR[i]
  mat[pA,pB] <- v; mat[pB,pA] <- v
}
```

```{r}
off <- mat[lower.tri(mat)]
message("\n--- wGRR summary ---")
message("  Pairs total   : ", length(off))
message("  wGRR > 0      : ", sum(off > 0))
message("  Mean wGRR     : ", round(mean(off), 4))
message("  Max  wGRR     : ", round(max(off),  4))
```
Now we turn the wGRR matrix into a distance matrix (distance = 1 - wGRR)

```{r}
dist_mat        <- 1 - mat
dist_mat[dist_mat < 0] <- 0                        # floating-point guard
diag(dist_mat)  <- 0                               # self-distance = 0

dist_obj <- as.dist(dist_mat)
```

```{r}
mds_result <- cmdscale(
d   = dist_obj,
k   = 2,
eig = TRUE,
add = TRUE)
```

```{r}
coords <- as.data.frame(mds_result$points)
colnames(coords) <- paste0("MDS", seq_len(2))
coords$plasmid_id <- rownames(mat)
coords <- coords[, c("plasmid_id", paste0("MDS", seq_len(2)))]  
```


```{r}
coords %>% ggplot(aes(x = MDS1, y = MDS2)) + geom_point(alpha = 0.5) + theme_bw()
```
```{r}
metadata <- read_tsv('data/metadata.tsv', col_names = c("accession", "species", "taxid", "length"))

metadata <- metadata %>% mutate(species_clean = stringr::word(species, 1, 2))
```

```{r}
coords <- merge(coords, metadata, by.x = "plasmid_id", by.y = "accession")
```

```{r}
coords %>% ggplot(aes(x = MDS1, y = MDS2, colour = species_clean)) + geom_point(alpha = 0.5) + theme_bw()

```

```{r}
library(gghighlight)

coords %>% ggplot(aes(x = MDS1, y = MDS2, colour = species_clean)) + 
  geom_point(alpha = 0.5) + 
  gghighlight() +
  facet_wrap(~ species_clean) + 
  theme_void() 


```
TODO:

Draw histograms of wGRR distributions in different species-species combinations 
