---
title: "Visualize wGRR plasmid networks for plant pathogen experiment"
output: html_document
date: "2026-03-10"
author: "Misha"
---

```{r}
library(dplyr)
library(readr)
library(stringr)
library(ggplot2)
library(tidyverse)
```

```{r}
dist_mat <- read.csv('../output/PP_wGRR_distance_matrix.csv', row.names = 1)
dist_obj <- as.dist(dist_mat)
```



```{r}
library(stringr)

metadata <- read_tsv('../data/metadata.tsv', col_names = c("accession", "species", "taxid", "length"))

metadata <- metadata %>% 
  mutate(species = gsub("Candidatus ", "", species, ignore.case = TRUE)) %>% 
  mutate(species_clean = stringr::word(species, 1, 2)) %>% 
  mutate(genus = stringr::word(species, 1, 1)) 
  
```

Merge metadata with the plasmid coordinates, plot with color for species:

I want to find plasmid pairs of plasmids from different genus, and then look for high wGRR values

```{r}
metadata_minimal <- metadata %>% select(accession, genus)

wgrr_pairs <- dist_mat %>%
  as.data.frame() %>%
  rownames_to_column("pA") %>%
  pivot_longer(-1, names_to = "pB", values_to = "distance") %>%
  rename(pA = 1) %>% 
  mutate(wGRR = 1 - distance) %>% 
  filter(pA != pB) %>%        # exclude self
  filter(pA < pB)         # deduplicate if symmetric


wgrr_pairs_species <- wgrr_pairs %>% 
  merge(metadata_minimal, by.x = "pA", by.y = "accession") %>% 
  mutate(genus_pA = genus) %>% 
  select(-genus) %>% 
  merge(metadata_minimal, by.x = "pB", by.y = "accession") %>% 
  mutate(genus_pB = genus) %>% 
  select(-genus) %>% 
  mutate(genus_pair = paste(genus_pA, genus_pB, sep = "-"))
```

```{r}
wgrr_pairs_species %>% group_by(genus_pair) %>% 
  summarise(mean = mean(wGRR),
            median = median(wGRR), 
            min = min(wGRR),
            max = max(wGRR))
```

```{r}
wgrr_pairs_species %>% 
  filter(genus_pA != genus_pB) %>% 
  filter(wGRR > 0.5)
```

Interesting Dickeya - Xanthomonas plasmid pair with 0.9 wGRR!


```{r}
library(ggraph)
library(tidygraph)
```
Alternative visualization in 2D space is a network (see Eduardo Rocha's papers). In these networks each plasmid becomes a node (a dot) and each wGRR connection (based on a threshold) is an edge (a line). It makes sense to only consider edges representing a certain level of similarity, e.g. wGRR > 0.5. 

```{r}
edges <- wgrr_pairs_species %>%
  filter(wGRR >= 0.5) %>%
  select(pA, pB, wGRR, genus_pA, genus_pB)

# Node table with species annotation
nodes <- data.frame(name = unique(c(edges$pA, edges$pB))) %>%
  left_join(
    wgrr_pairs_species %>% select(pA, genus_pA) %>% distinct() %>% rename(name = pA, species = genus_pA),
    by = "name"
  )

g <- tbl_graph(nodes = nodes, edges = edges, directed = FALSE, node_key = "name")
```

Draw the network:

```{r}
ggraph(g, layout = "fr") +  # Fruchterman-Reingold, good for clusters
  geom_edge_link(aes(alpha = wGRR), colour = "grey60") +
  geom_node_point(aes(colour = species), size = 1) +
  # geom_node_text(aes(label = name), repel = TRUE, size = 2.5) +
  #scale_edge_width(range = c(0.3, 2)) +
  scale_edge_alpha(range = c(0.3, 0.9)) +
  #scale_colour_brewer(palette = "Set2") +
  theme_graph() +
  labs(title = "Plasmid relatedness network (wGRR ≥ 0.5)",
       colour = "Host species", edge_width = "wGRR") + theme_void()
```
Highlight cross-species edges in the network:

```{r}
g <- g %>%
  activate(edges) %>%
  mutate(cross_species = genus_pA != genus_pB)

ggraph(g, layout = "fr") +
  geom_edge_link(aes(colour = cross_species, width = wGRR, alpha = wGRR)) +
  geom_node_point(aes(colour = species), size = 2) +
  scale_edge_colour_manual(values = c("FALSE" = "grey80", "TRUE" = "firebrick")) +
  scale_edge_width(range = c(0.3, 2)) +
  scale_edge_alpha(range = c(0.2, 0.9)) +
  theme_graph() +
  labs(title = "Cross-species plasmid transfer network",
       edge_colour = "Cross-species")
```


Repeat but only include cross species interactions in building the network in the first place

```{r}
# keep only cross-species edges
edges_cross <- wgrr_pairs_species %>%
  filter(wGRR >= 0.5, genus_pA != genus_pB) %>%
  select(pA, pB, wGRR, genus_pA, genus_pB)

# nodes automatically limited to only those in cross-species edges
nodes_cross <- data.frame(name = unique(c(edges_cross$pA, edges_cross$pB))) %>%
  left_join(
    wgrr_pairs_species %>% select(pA, genus_pA) %>% distinct() %>% rename(name = pA, species = genus_pA),
    by = "name"
  )

g_cross <- tbl_graph(nodes = nodes_cross, edges = edges_cross, directed = FALSE, node_key = "name")

```

```{r}
ggraph(g_cross, layout = "fr") +
  geom_edge_link(aes(width = wGRR), colour = "firebrick") +
  geom_node_point(aes(colour = species), size = 1) +
  # scale_colour_brewer(palette = "Set2") +
  scale_edge_width(range = c(0.3, 2)) +
  theme_graph() +
  labs(title = "Cross-species plasmid transfer network (wGRR ≥ 0.5)",
       colour = "Host species")
```
Pretty cool!

Each sub graph thats not connected to others is called a 'component'. We can add that information and then pull out the particular accession names in each component.

```{r}
# add component membership
g_cross <- g_cross %>%
  activate(nodes) %>%
  mutate(component = group_components())

# summarise each component: size, n species, mean wGRR
component_summary <- g_cross %>%
  activate(nodes) %>%
  as_tibble() %>%
  group_by(component) %>%
  summarise(
    n_plasmids = n(),
    n_species = n_distinct(species),
    .groups = "drop"
  ) %>%
  arrange(desc(n_species), desc(n_plasmids))
```

I'm curious about component 8 in this case, which contains 5 plasmids from 4 different species!

```{r}
target_component <- 7 # replace with your component of interest

plasmids_of_interest <- g_cross %>%
  activate(nodes) %>%
  as_tibble() %>%
  filter(component == target_component) 

cat(plasmids_of_interest %>% pull(name))
```



Final visualization option: UMAP

```{r}
library(uwot)
library(ggplot2)

dist_mat_object <- dist(dist_mat)

# run UMAP on the distance matrix
umap_res <- umap(dist_mat, n_neighbors = 10, min_dist = 2)

# build a plottable dataframe
umap_df <- data.frame(
  UMAP1 = umap_res[, 1],
  UMAP2 = umap_res[, 2],
  plasmid = rownames(dist_mat)
) %>%
  left_join(
    wgrr_pairs_species %>% select(pA, species_pA) %>% distinct() %>% rename(plasmid = pA, species = species_pA),
    by = "plasmid"
  )

ggplot(umap_df, aes(UMAP1, UMAP2, colour = species)) +
  geom_point(size = 1.5, alpha = 0.7) +
  theme_void() +
  labs(colour = "Host species")
```



