---
title: "Visualize wGRR plasmid networks"
output: html_document
date: "2026-03-10"
author: "Misha"
---

```{r}
library(dplyr)
library(readr)
library(stringr)
library(ggplot2)
library(tidyverse)
```

```{r}
dist_mat <- read.csv('../output/xanthomonas_wGRR_distance_matrix.csv', row.names = 1)
dist_obj <- as.dist(dist_mat)
```

Visualizing a distance matrix can be done in many ways. Here we explore a dimensionality reduction method (MDS) to visualize the plasmids in 2D space (somewhat similar to PCA):

```{r}
mds_result <- cmdscale(
  d   = dist_obj,
  k   = 2,
  eig = TRUE,
  add = TRUE
  )
```

Just like a PCA we can find the variance explained by the two axes.

```{r}
variance_explained <- mds_result$eig / sum(mds_result$eig[mds_result$eig > 0]) * 100
cat(round(variance_explained[1:2], 1), "%")
```
Oof that's very low! Keep that in mind while looking at the following scatter plots.

To plot a scatter plot first we get the coordinates from the mds_results

```{r}
coords <- as.data.frame(mds_result$points)
colnames(coords) <- paste0("MDS", seq_len(2))
coords$plasmid_id <- rownames(dist_mat)
coords <- coords[, c("plasmid_id", paste0("MDS", seq_len(2)))]  
```

Plot plasmids in MDS space:

```{r}
coords %>% ggplot(aes(x = MDS1, y = MDS2)) + geom_point(alpha = 0.5) + theme_bw()
```
```{r}
metadata <- read_tsv('../data/xanthomonas_metadata.tsv', col_names = c("accession", "species", "taxid", "length"))

metadata <- metadata %>% mutate(species_clean = stringr::word(species, 1, 2))
```
Merge metadata with the plasmid coordinates, plot with color for species:

```{r}
coords <- merge(coords, metadata, by.x = "plasmid_id", by.y = "accession")
```

```{r}
coords %>% ggplot(aes(x = MDS1, y = MDS2, colour = species_clean)) + geom_point(alpha = 0.5) + theme_bw()
```
This is quite cool! It shows different clusters of plasmids, with different colors within clusters. This could indicate plasmid (parts) being shared between different species, however, we need to keep in mind the low percentage of variance explained with the MDA.

```{r}
coords %>% filter(MDS1 > 0.5) %>% filter(MDS2 > 0) %>% 
  ggplot(aes(x = MDS1, y = MDS2, colour = species_clean, size = length)) + 
  geom_point(alpha = 1) + 
  scale_colour_brewer(palette = "Paired") +
  theme_void()
 

```

```{r}
library(gghighlight)

coords %>% ggplot(aes(x = MDS1, y = MDS2, colour = species_clean)) + 
  geom_point(alpha = 0.5) + 
  gghighlight() +
  facet_wrap(~ species_clean) + 
  theme_void() 
```
I want to find plasmid pairs of plasmids from different species, and then look for high wGRR values

```{r}
metadata_minimal <- metadata %>% select(accession, species_clean)

wgrr_pairs <- dist_mat %>%
  as.data.frame() %>%
  rownames_to_column("pA") %>%
  pivot_longer(-1, names_to = "pB", values_to = "distance") %>%
  rename(pA = 1) %>% 
  mutate(wGRR = 1 - distance) %>% 
  filter(pA != pB) %>%        # exclude self
  filter(pA < pB)         # deduplicate if symmetric


wgrr_pairs_species <- wgrr_pairs %>% 
  merge(metadata_minimal, by.x = "pA", by.y = "accession") %>% 
  mutate(species_pA = species_clean) %>% 
  select(-species_clean) %>% 
  merge(metadata_minimal, by.x = "pB", by.y = "accession") %>% 
  mutate(species_pB = species_clean) %>% 
  select(-species_clean) %>% 
  mutate(species_pair = paste(species_pA, species_pB, sep = "-"))
```

```{r}
wgrr_pairs_species %>% group_by(species_pair) %>% 
  summarise(mean = mean(wGRR),
            median = median(wGRR), 
            min = min(wGRR),
            max = max(wGRR))
```

```{r}
wgrr_pairs_species %>% 
  filter(species_pA != species_pB) %>% 
  filter(wGRR > 0.95)
```

The final one is a really interesting result. It should cross-species plasmid pairs with very high wGRR values.


```{r}
library(ggraph)
library(tidygraph)
```
Alternative visualization in 2D space is a network (see Eduardo Rocha's papers). In these networks each plasmid becomes a node (a dot) and each wGRR connection (based on a threshold) is an edge (a line). It makes sense to only consider edges representing a certain level of similarity, e.g. wGRR > 0.5. 

```{r}
edges <- wgrr_pairs_species %>%
  filter(wGRR >= 0.3) %>%
  select(pA, pB, wGRR, species_pA, species_pB)

# Node table with species annotation
nodes <- data.frame(name = unique(c(edges$pA, edges$pB))) %>%
  left_join(
    wgrr_pairs_species %>% select(pA, species_pA) %>% distinct() %>% rename(name = pA, species = species_pA),
    by = "name"
  )

g <- tbl_graph(nodes = nodes, edges = edges, directed = FALSE, node_key = "name")
```

Draw the network:

```{r}
ggraph(g, layout = "fr") +  # Fruchterman-Reingold, good for clusters
  geom_edge_link(aes(alpha = wGRR), colour = "grey60") +
  geom_node_point(aes(colour = species), size = 1) +
  # geom_node_text(aes(label = name), repel = TRUE, size = 2.5) +
  #scale_edge_width(range = c(0.3, 2)) +
  scale_edge_alpha(range = c(0.3, 0.9)) +
  #scale_colour_brewer(palette = "Set2") +
  theme_graph() +
  labs(title = "Plasmid relatedness network (wGRR ≥ 0.4)",
       colour = "Host species", edge_width = "wGRR") + theme_void()
```
Highlight cross-species edges in the network:

```{r}
g <- g %>%
  activate(edges) %>%
  mutate(cross_species = species_pA != species_pB)

ggraph(g, layout = "fr") +
  geom_edge_link(aes(colour = cross_species, width = wGRR, alpha = wGRR)) +
  geom_node_point(aes(colour = species), size = 2) +
  scale_edge_colour_manual(values = c("FALSE" = "grey80", "TRUE" = "firebrick")) +
  scale_edge_width(range = c(0.3, 2)) +
  scale_edge_alpha(range = c(0.2, 0.9)) +
  theme_graph() +
  labs(title = "Cross-species plasmid transfer network",
       edge_colour = "Cross-species")
```
Repeat but only include cross species interactions in building the network in the first place

```{r}
# keep only cross-species edges
edges_cross <- wgrr_pairs_species %>%
  filter(wGRR >= 0.8, species_pA != species_pB) %>%
  select(pA, pB, wGRR, species_pA, species_pB)

# nodes automatically limited to only those in cross-species edges
nodes_cross <- data.frame(name = unique(c(edges_cross$pA, edges_cross$pB))) %>%
  left_join(
    wgrr_pairs_species %>% select(pA, species_pA) %>% distinct() %>% rename(name = pA, species = species_pA),
    by = "name"
  )

g_cross <- tbl_graph(nodes = nodes_cross, edges = edges_cross, directed = FALSE, node_key = "name")
```

```{r}
ggraph(g_cross, layout = "fr") +
  geom_edge_link(colour = "firebrick") +
  geom_node_point(aes(colour = species), size = 1) +
  # scale_colour_brewer(palette = "Set2") +
  theme_graph() +
  labs(title = "Cross-species plasmid transfer network (wGRR ≥ 0.8)",
       colour = "Host species")
```
Pretty cool!

Each sub graph thats not connected to others is called a 'component'. We can add that information and then pull out the particular accession names in each component.

```{r}
# add component membership
g_cross <- g_cross %>%
  activate(nodes) %>%
  mutate(component = group_components())

# summarise each component: size, n species, mean wGRR
component_summary <- g_cross %>%
  activate(nodes) %>%
  as_tibble() %>%
  group_by(component) %>%
  summarise(
    n_plasmids = n(),
    n_species = n_distinct(species),
    .groups = "drop"
  ) %>%
  arrange(desc(n_species), desc(n_plasmids))
```

I'm curious about component 8 in this case, which contains 5 plasmids from 4 different species!

```{r}
target_component <- 2  # replace with your component of interest

plasmids_of_interest <- g_cross %>%
  activate(nodes) %>%
  as_tibble() %>%
  filter(component == target_component) 

cat(plasmids_of_interest %>% pull(name))
```



Final visualization option: UMAP

```{r}
library(uwot)
library(ggplot2)

dist_mat_object <- dist(dist_mat)

# run UMAP on the distance matrix
umap_res <- umap(dist_mat, n_neighbors = 10, min_dist = 2)

# build a plottable dataframe
umap_df <- data.frame(
  UMAP1 = umap_res[, 1],
  UMAP2 = umap_res[, 2],
  plasmid = rownames(dist_mat)
) %>%
  left_join(
    wgrr_pairs_species %>% select(pA, species_pA) %>% distinct() %>% rename(plasmid = pA, species = species_pA),
    by = "plasmid"
  )

ggplot(umap_df, aes(UMAP1, UMAP2, colour = species)) +
  geom_point(size = 1.5, alpha = 0.7) +
  theme_void() +
  labs(colour = "Host species")
```



